---
title: "Computing a PGS using RápidoPGS-single and GWAS catalog"
author: "Guillermo Reales"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing a PGS using RápidoPGS-single and GWAS catalog}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}

---

# Introduction

RápidoPGS is a tool to quickly compute polygenic scores (PGS) from GWAS summary statistics datasets from either case-control (eg. asthma) or quantitative traits (eg. height and BMI).

Input GWAS summary statistics datasets should have a minimum set of columns, 
with these names, but in any order:

* **CHR**: Chromosome 
* **BP**: Base position (in GRCh37/hg19 or GRCh38/hg38). If using hg38, use build = "hg38" in parameters.
* **SNPID**: rsids, or SNP identifiers. If not available, they can be anything (eg. CHR_BP).
* **REF**: Reference, or non-effect allele
* **ALT**: Alternative, or effect allele, the one \eqn{\beta} refers to
* **BETA**: $\beta$ (or log(OR)), or effect sizes.
* **SE**: Standard error of $\beta$
* **P**: P-values for the association test

Also, if doing a PGS on a quantitative trait GWAS dataset, your file must include this:
* **ALT_FREQ**: Minor/ALT allele frequency in the tested population, or in a close population from a reference panel. 


# Installation of RápidoPGS

Current RápidoPGS development version (v2.1.0) is available on GitHub, so we can install it using the following code (Note: If you don't have `remotes` installed, please install it first:

```{r eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("GRealesM/RapidoPGS")
```

Note that previous version of RápidoPGS (v1.0.2) is available on CRAN, but we changed important things in the package, including functions for RápidoPGS-multi approach, so until we publish the current version we highly recommend downloading the current development version, to which this vignette refers to.


# Setup

Once installed, let's load it. This will automatically load all required dependencies.

```{r message=FALSE, warning = FALSE}
library(RapidoPGS)
```

# Downloading data from GWAS catalog

[GWAS catalog](https://www.ebi.ac.uk/gwas/) is an outstanding GWAS summary statistics data source, as it not only puts together tons of public datasets, but it also uses a semi-automatic pipeline to apply quality control and liftover (a.k.a. harmonise) those datasets, thus helping overcome the problem of having GWAS sumstats in so many different formats and builds.

In this vignette, we'll use GWAS catalog to obtain an example dataset. For this vignette we'll use a Breast cancer (BRCA) dataset from [Michailidou et al., 2017](https://www.nature.com/articles/nature24284), which is one that we used in our manuscript. This GWAS was performed on 119078 controls and 137045 cases of Breast cancer.

It's a big file, and may take a while to download, so here we will show the command to download, but actually cheat and load a subset of data already loaded.

Note that in this particular case we chose a study that contained a single dataset. In other cases there may be more than one. In that case `gwascat.download` will prompt you with the list of available options, prompting their accession numbers, and asking you to choose a file by its number in the list, if running interactively. If you want to run `gwascat.download` automatically, and you know the number already, you can supply it using `filenum` argument, so R won't bug you. 

We use it's PubMed ID (29059683). We could download only harmonised GWAS catalog columns (thus having a lighter file) but I don't advise this because sometimes we may need columns from the original file. We then set `hm_only = FALSE` to get all columns.

```{r eval =FALSE}
ds <- gwascat.download(29059683, hm_only = FALSE)
```


Now for the cheat: for illustrative purposes, I took a random subset from this file including 100,000 SNPs from this file, which we'll use in this tutorial. Bear in mind that the final PGS won't be a *valid* model since we randomly removed most SNPs. It will serve for our teaching purposes, though! ;)


```{r}
ds <- michailidou
```


# Applying quality control to the dataset

The first thing to do is to check what our file looks like. RápidoPGS can handle NAs in crucial columns, so don't worry too much about that (unless you have all NA for crucial columns, of course!).

A note of caution when dealing with GWAS catalog files, though: since they use a semi-automated pipeline, it is possible that even some columns are present, they are *empty*, so be careful with that! 

Also, RápidoPGS requires allele frequencies, so it's possible that you need to compute it from an external reference panel (eg. 1000 Genomes Phase III). We don't cover that step in this vignette, but we might write instructions explaining how to do it in the future.

Let's take a look at the file!

```{r}
summary(ds)
```
In this case, we don't have any missing data, which is fantastic. We'll now prepare the dataset for RápidoPGS. We need to rename columns to something RápidoPGS can understand, and this is easy to do with data.table.

```{r}
setnames(ds, old = c("hm_rsid", "hm_chrom", "hm_pos", "hm_other_allele", "hm_effect_allele", "hm_beta", "hm_effect_allele_frequency", "standard_error", "p_value"), new=c("SNPID","CHR", "BP","REF", "ALT", "BETA", "ALT_FREQ", "SE", "P"))
```

Now another thing to consider: RápidoPGS use only autosomes, so remove X or Y chromosomes at this step.

```{r}
ds <- ds[CHR != "X",]
```


# Computing PGS using RápidoPGS-single

Now we have applied QC to our dataset, we can now compute our PGS!
We can easily do that applying the next command, but bear in mind that, since we used harmonised columns, the build is **hg38**, so we must tell RápidoPGS about that.

In this new version, we don't need sample size for case-control datasets.
Note that if this was a quantitative trait dataset, we should inform total number of individuals (N parameter). Also, if our dataset had columns reporting the number of individuals for each SNP, we can replace N by a string specifying the column (eg. N = "sample_size"). By doing so, RápidoPGS-single will extract this information directly from the columns.

Let's get our hands dirty! Let's compute first a *full* RápidoPGS-single model. 
Since we used harmonised coordinates from GWAS catalog, our file is in build hg38, and we need to specify it to RápidoPGS (default is hg19). 

**Advanced use note**: You may want to filter by and align your dataset to an external reference. You can do that with RápidoPGS, too, by specifying the path of your reference file at `reference` argument. This reference file should have five columns (CHR, BP, REF, ALT, and SNPID) and be in the **same build** as our summary statistics dataset. RápidoPGS currently supports **hg19** and **hg38** builds.

```{r}
full_PGS <- rapidopgs_single(ds, trait = "cc", build = "hg38")
```

Well, that was *rápido*! With increasingly big datasets, it will take a bit longer, of course.

Let's take a look.

```{r}
head(full_PGS)
```

We see three new columns: **ld.block** corresponds to the [ld-detect](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4731402/) LD block number assigned to the SNP (see manuscript for more details of where this comes from), **ppi** is the posterior probability that the SNP is causal, and **weight** is the weighted effect size (BETA * ppi) - and the column we're interested in.


## Applying different thresholds

Imagine that we want a small portable PGS.  We could use a threshold (eg. keep only SNPs with ppi larger than $1e^{-4}$, a reasonable threshold) or keep the top SNPs with largest weights (in either way).

We can do that using RápidoPGS, let's see how by using a $1e^{-4}$ threshold.

For accuracy reasons, we recommend recomputing the PGS on just these SNPs after the threshold was applied, so `recalc = TRUE` by default. 

```{r}
PGS_1e4 <- rapidopgs_single(ds, trait ="cc", build = "hg38", filt_threshold = 1e-4)
head(PGS_1e4)
```

You can omit recalculation by setting that argument to `FALSE`

```{r}
PGS_1e4_norecalc <- rapidopgs_single(ds, trait ="cc", build = "hg38", filt_threshold = 1e-4, recalc = FALSE)
head(PGS_1e4_norecalc)
```

And what if we want the *top 10* SNPs? Just change the `filt_threshold` argument. If it's larger than 1, RápidoPGS will understand that you want a top list, rather than a thresholded result.

```{r}
PGS_top10 <- rapidopgs_single(ds, trait ="cc", build = "hg38", filt_threshold = 10)
head(PGS_top10)
```


# Conclusion

So those are examples of basic *RápidoPGS-single* usage! 

Drop us a line if you encounter problems, and we'll be happy to help.

Good luck!


Guillermo
